Reconstructing the last known movements of one of Nature's giants

### 2017-06-21 manuscript/blue-whale-ms.tex
Minor typo in acknowlegments
Fix symbols

\usepackage[osf]{mathpazo} % palatino
\usepackage{ms}            % load the template
\usepackage[round]{natbib} % author-year citations
\usepackage{graphicx}
\usepackage{parskip}      
\pagenumbering{arabic}    
\linespread{1.66}

@@

Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
It is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert (Best and Schell 1996). 
We sampled keratin from the blue whale’s baleen plate at 1cm intervals and used continuous-flow isotope ratio mass spectrometry to extract $delta^13C$ and $delta^15N$ isotope concentrations. 
Given the date of stranding (25th March 1891), and estimated baleen growth rates of $13.5 cm y^-1$ (see Supplemental Methods), we reconstructed a timeline for $delta^13C$ and $delta^15N$ fluctuations in the baleen over the last six years of whale’s life (autumn 1885 - spring 1891). The $delta^13C$ and $delta^15N$ profiles show two distinct phases in the whale's behaviour. 
In behavioural phase one (month 1885 to May 1889), we find strong cyclical fluctuations in both $delta^13C$ and $delta^15N$ values of an approximately yearly periodicity (Fig. 1). $delta^13C$ and $delta^15N$ exhibit strong negative covariance. 
In behavioural phase two (June 1889 to March 1891) we see a sharp uptick in $delta^13C$, followed by high and relatively constant $delta^13C$ values. 
This is associated with an overall reduction in $delta^15N$ values and a weakening of the periodic $delta^15N$ fluctuations seen in behavioural phase one. There is also no covariance between $delta^13C$ and $delta^15N$ (Fig. 1).

@@

To interpret these isotopic signatures, we simulated time series of $delta^13C$ and $delta^15N$ values expected under differing movement behaviours, by coupling a novel agent-based model of whale movements (see Supplementary Methods) to models of global phytoplankton $delta^13C$ and $delta^15N$ values (Schmittner and Somes; Magozzi et al 2017). 
We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth and phytoplankton concentration (as a proxy for zooplankton food availability). 
At each time point and location in the simulations, we extracted phytoplankton $delta^13C$ and $delta^15N$ values from the models (Schmittner and Somes, Magozzi et al 2017) resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas. 
We then determined which of these simulated profiles were biogeochemically consistent with the isotopic profile of the blue whale’s baleen (Fig. 1) and use this to interpret the results.

During behavioural phase one, the whale is foraging at high northern latitudes, likely off the coast of Norway. 
The pronounced cyclical variations in $delta^13C$ values are characteristic of the isotopic expression of the spring phytoplankton bloom in northern waters (Magozzi et al 2017). 
This interpretation is also consistent with the relatively negative median $delta^13C$ values (Fig. 1). 
The periodicity of $delta^13C$ and $delta^15N$ values is explained by short distance seasonal migrations, with the whale overwintering off the coast of Ireland. 
The relatively positive and invariant $delta^13C$ values seen during behavioural phase two are found in subtropical areas, particularly off the coast of the Azores, suggesting the whale migrated rapidly to southerly latitudes in early 1889, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods; Busquets-Vass 2017 ?silva paper). 
The whale remained in these warm waters for over a year and was returning to her northern feeding grounds when she became stranded.

Movement to warm waters alone, however, cannot explain the dramatic increase in $delta^13C$ values at the start of behavioural phase two. 
We believe that this reflects the physiological demands of pregnancy, as similar increases in $delta^13C$ values have been seen in pregnant South American sea lions (Cardona et al 2017). 
The gradual increase in $delta^15N$ values to the autumn of 1890 is consistent with the isotopic expression associated with lactation (Cardona et al 2017). Blue whales have a 10-12 month gestation period, and calves are weaned after 6-7 months (Wilson  Mittermeier 2014). 
The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the known reproductive ecology of blue whales. 
This is the first quantitative evidence of historical movement patterns in blue whales, and reveals a fascinating snapshot into the life of an iconic specimen, where we previously only knew the date and location of its death.

@@

\subsection{Stable isotope extractions from baleen}
Baleen was collected from the Natural History Museum, London (specimen BMNH.1892.3.1.1). 
The baleen plate was cleaned with ethanol to remove surface contaminants. 
$~$1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
78 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section which contains the most recent tissue. 
Baleen grows at a constant rate, so the samples are equally spaced through time (Best and Schell 1996). Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the National Oceanographic Centre (Southampton, UK), using a Vario Isotope select elemental analyser, coupled with Isoprime 100 isotope mass spectrometer. 
Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
C:N ratios for samples ranged from 3.28$‰$ to 3.72$ppm$, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples (Hobson  Koch 1998).

\subsection{Time calibrating stable isotope profiles}
Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates  (Hobson and Schell, 1987, Busquets-Vass et al 2017). 
Clear periodicity was evident in $delta^15N$ values across the entire baleen plate, and in $delta^13C$ values in the most distal (oldest) 65cm of the plate. 
We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis (Cardona et al 2017), revealing a consistent growth rate of 13.5cm $y^-1$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cm y^-1$ estimated for six blue whales from the Californian Pacific (Busquets-Vass et al 2017).  
Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. Cross-correlation analysis (ref?) demonstrated a strong negative covariance between $delta^13C$ and $delta^15N$ values within behavioural phase one (oldest 60cm of the baleen plate), but no relationship between $delta^13C$ and $delta^15N$ values during behavioural phase two.

\subsection{Baseline isotope comparisons}
Isotope-enabled biogeochemical ocean models (Schmittner and Somes, Magozzi et al 2017) were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds. Annual average $delta^15N$ POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{circ}$ resolution biogeochemical model. $delta^13C$ POM values were simulated at 1$^{circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model (Magozzi et al 2017, Yool et al 2013). 
Simulated  $delta^15N$ POM values are relatively positive in the northeast Atlantic north of c. 60$^{circ}$N, and relatively negative in the central and southern North Atlantic. Annual average $delta^13C$ POM values largely vary with latitude, with more negative values in more northerly regions. In the central North Atlantic, $delta^13C$ POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig S3). 
The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature (Laws, Magozzi et al 2017). 
Thus temporal variations in $delta^13C$ POM values were superimposed on latitudinal gradients. 
The scale and nature of temporal variation in $delta^13C$ POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in $delta^13C$ POM values linked to strongly seasonal phytoplankton growth dynamics. 
We therefore used monthly simulated $delta^13C$ POM values to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours.

@@

In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and extent (1$^{circ}$ or 2$^{circ}$) of movement are all influenced by the following. 
(i) Behavioural state (migrating north, migrating south, foraging, or nursing). 
Northerly migrations were possible only in spring, and southerly migrations in autumn. Foraging was possible at anytime of year, and was triggered when whales encountered high concentrations of plankton. Nursing occurred at temperatures between 18 and 25$^{circ}$C only. 
(ii) Sea surface temperature ($^{circ}$C, Yool et al 2013). 
When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{circ}$C); whereas whales migrating south sought warmer waters. 
(iii) Water depth (m; what reference is this collected from?). 
Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
(iv) Phytoplankton concentration (mmol N m-3, for combined diatom and non-diatom communities, Yool et al 2013). 
This was included as a proxy for zooplankton food availability. 
Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.

We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores; Fig SX; McDonald et al 2006; Reilly et al 2008; sigurjonsson 1995); (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes. 
Various movement rules were implemented to avoid simulations ending up on land, or becoming “stuck” in the relatively shallow waters of the Bay of Biscay or the Strait of Dover. 
For a full description of the model and its parameters see Supplementary Methods. We simulated whale movements XX times for each scenario and residency hotspot. 
For each movement simulation we extracted phytoplankton $delta^13C$ and $delta^15N$ values at each time point and location from the models described above (Magozzi, Somes). 
We then compared the simulated stable isotope profiles (Fig. S4) to the profile of the blue whale (Fig. 1).

@@

\section{Acknowledgements}
% adjectives not allowed!
This work was funded by the British Ecological Society (grant: 5771/6815). 
We thank XXX for comments on drafts.
We thank CJ Somes for providing  δ13C POM data.

>>>>

\usepackage[osf]{mathpazo} % palatino
\usepackage{ms}            % load the template
\usepackage[round]{natbib} % author-year citations
\usepackage{graphicx}
\usepackage{parskip} 
\usepackage{textcomp} % for parts per mille symbol     
\pagenumbering{arabic}    
\linespread{1.66}

@@

Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
It is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert (Best and Schell 1996). 
We sampled keratin from the blue whale's baleen plate at 1cm intervals and used continuous-flow isotope ratio mass spectrometry to extract $\delta^{13}$C and $\delta^{15}$N isotope concentrations. 
Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplemental Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the last six years of whale's life (autumn 1885 - spring 1891). The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
In behavioural phase one (month 1885 to May 1889), we find strong cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values of an approximately yearly periodicity (Fig. 1). $\delta^{13}$C and $\delta^{15}$N exhibit strong negative covariance. 
In behavioural phase two (June 1889 to March 1891) we see a sharp uptick in $\delta^{13}$C, followed by high and relatively constant $\delta^{13}$C values. 
This is associated with an overall reduction in $\delta^{15}$N values and a weakening of the periodic $\delta^{15}$N fluctuations seen in behavioural phase one. There is also no covariance between $\delta^{13}$C and $\delta^{15}$N (Fig. 1).

@@

To interpret these isotopic signatures, we simulated time series of $\delta^{13}$C and $\delta^{15}$N values expected under differing movement behaviours, by coupling a novel agent-based model of whale movements (see Supplementary Methods) to models of global phytoplankton $\delta^{13}$C and $\delta^{15}$N values (Schmittner and Somes; Magozzi et al 2017). 
We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth and phytoplankton concentration (as a proxy for zooplankton food availability). 
At each time point and location in the simulations, we extracted phytoplankton $\delta^{13}$C and $\delta^{15}$N values from the models (Schmittner and Somes, Magozzi et al 2017) resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas. 
We then determined which of these simulated profiles were biogeochemically consistent with the isotopic profile of the blue whale’s baleen (Fig. 1) and use this to interpret the results.

During behavioural phase one, the whale is foraging at high northern latitudes, likely off the coast of Norway. 
The pronounced cyclical variations in $\delta^{13}$C values are characteristic of the isotopic expression of the spring phytoplankton bloom in northern waters (Magozzi et al 2017). 
This interpretation is also consistent with the relatively negative median $\delta^{13}$C values (Fig. 1). 
The periodicity of $\delta^{13}$C and $\delta^{15}$N values is explained by short distance seasonal migrations, with the whale overwintering off the coast of Ireland. 
The relatively positive and invariant $\delta^{13}$C values seen during behavioural phase two are found in subtropical areas, particularly off the coast of the Azores, suggesting the whale migrated rapidly to southerly latitudes in early 1889, but continued feeding (other mysticete whales fast during migrations, but blue whales are too large to stop feeding for long periods; Busquets-Vass 2017 ?silva paper). 
The whale remained in these warm waters for over a year and was returning to her northern feeding grounds when she became stranded.

Movement to warm waters alone, however, cannot explain the dramatic increase in $\delta^{13}$C values at the start of behavioural phase two. 
We believe that this reflects the physiological demands of pregnancy, as similar increases in $\delta^{13}$C values have been seen in pregnant South American sea lions (Cardona et al 2017). 
The gradual increase in $\delta^{15}$N values to the autumn of 1890 is consistent with the isotopic expression associated with lactation (Cardona et al 2017). Blue whales have a 10-12 month gestation period, and calves are weaned after 6-7 months (Wilson  Mittermeier 2014). 
The proposed timescale of pregnancy around early spring 1889, birth in spring 1890, lactation through summer 1890 and return to northern feeding grounds in early 1891 is therefore fully consistent with the known reproductive ecology of blue whales. 
This is the first quantitative evidence of historical movement patterns in blue whales, and reveals a fascinating snapshot into the life of an iconic specimen, where we previously only knew the date and location of its death.

@@

\subsection{Stable isotope extractions from baleen}
Baleen was collected from the Natural History Museum, London (specimen BMNH.1892.3.1.1). 
The baleen plate was cleaned with ethanol to remove surface contaminants. 
{\raise.17ex\hbox{$\scriptstyle\sim$}}1mg samples of keratin powder were then collected from the plate using a hand-held drill and grinding bit. 
78 samples were taken at 1cm intervals, 0.5cm from the outer edge of the plate, starting at the proximal section which contains the most recent tissue. 
Baleen grows at a constant rate, so the samples are equally spaced through time (Best and Schell 1996). Carbon and nitrogen isotope analysis was performed simultaneously via continuous-flow isotope ratio mass spectrometry at the National Oceanographic Centre (Southampton, UK), using a Vario Isotope select elemental analyser, coupled with Isoprime 100 isotope mass spectrometer. 
Replicates using internal laboratory standards (L-glutamic acid (C), Glutamic acid (CT standard), acetanilide and protein standard OAS) were used for quality control and calibration. 
C:N ratios for samples ranged from 3.28\text{\textperthousand} to 3.72\text{\textperthousand}, well within the acceptable theoretical range for pure keratin ($3.4\pm0.5$) allowing for comparison among samples (Hobson  Koch 1998).

\subsection{Time calibrating stable isotope profiles}
Seasonal migrations across isotopic gradients induce cyclical variations in the isotopic composition of baleen, the distance between cycles reflecting growth rates  (Hobson and Schell, 1987, Busquets-Vass et al 2017). 
Clear periodicity was evident in $\delta^{15}$N values across the entire baleen plate, and in $\delta^{13}$C values in the most distal (oldest) 65cm of the plate. 
We calculated isotopic periodicity within the baleen sample using Fourier Transform analysis (Cardona et al 2017), revealing a consistent growth rate of 13.5cm $y^{-1}$ which is remarkably similar to the mean isotope-derived baleen growth rates of $15.5 \pm 2.2cm y{^-1}$ estimated for six blue whales from the Californian Pacific (Busquets-Vass et al 2017).  
Therefore we dated the youngest baleen sample as 1st March 1891, 24 days prior to the stranding date, 25th March 1891. Cross-correlation analysis (ref?) demonstrated a strong negative covariance between $\delta^{13}$C and $\delta^{15}$N values within behavioural phase one (oldest 60cm of the baleen plate), but no relationship between $\delta^{13}$C and $\delta^{15}$N values during behavioural phase two.

\subsection{Baseline isotope comparisons}
Isotope-enabled biogeochemical ocean models (Schmittner and Somes, Magozzi et al 2017) were used to characterize the isotopic composition of phytoplankton expected in different potential foraging grounds. Annual average $\delta^{15}$N POM (particulate organic matter) values were provided by Somes (pers.comm) based on a 5$^{\circ}$ resolution biogeochemical model. $\delta^{13}$C POM values were simulated at 1$^{\circ}$ and monthly resolution using an isotopic extension to the NEMO-MEDUSA ocean biogeochemical model (Magozzi et al 2017, Yool et al 2013). 
Simulated  $\delta^{15}$N POM values are relatively positive in the northeast Atlantic north of c. 60$^{\circ}$N, and relatively negative in the central and southern North Atlantic. Annual average $\delta^{13}$C POM values largely vary with latitude, with more negative values in more northerly regions. In the central North Atlantic, $\delta^{13}$C POM values are relatively positive in the west, reflecting warm gulf stream waters (Fig S3). 
The isotopic composition of carbon in phytoplankton also varies through seasons as isotopic fractionation of carbon during photosynthesis is strongly influenced by sea surface temperature (Laws, Magozzi et al 2017). 
Thus temporal variations in $\delta^{13}$C POM values were superimposed on latitudinal gradients. 
The scale and nature of temporal variation in $\delta^{13}$C POM values also varies with latitude, with higher latitude seas showing greater intra-annual variation in $\delta^{13}$C POM values linked to strongly seasonal phytoplankton growth dynamics. 
We therefore used monthly simulated $\delta^{13}$C POM values to simulate the isotopic expression of phytoplankton expected to be encountered by whales exhibiting differing movement behaviours.

@@

In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and extent (1$^{\circ}$ or 2$^{\circ}$) of movement are all influenced by the following. 
(i) Behavioural state (migrating north, migrating south, foraging, or nursing). 
Northerly migrations were possible only in spring, and southerly migrations in autumn. Foraging was possible at anytime of year, and was triggered when whales encountered high concentrations of plankton. Nursing occurred at temperatures between 18 and 25$^{\circ}$C only. 
(ii) Sea surface temperature ($^{\circ}$C, Yool et al 2013). 
When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters. 
(iii) Water depth (m; what reference is this collected from?). 
Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
(iv) Phytoplankton concentration (mmol N $m^{-3}$, for combined diatom and non-diatom communities, Yool et al 2013). 
This was included as a proxy for zooplankton food availability. 
Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.

We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores; Fig SX; McDonald et al 2006; Reilly et al 2008; sigurjonsson 1995); (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes. 
Various movement rules were implemented to avoid simulations ending up on land, or becoming `stuck' in the relatively shallow waters of the Bay of Biscay or the Strait of Dover. 
For a full description of the model and its parameters see Supplementary Methods. We simulated whale movements XX times for each scenario and residency hotspot. 
For each movement simulation we extracted phytoplankton $\delta^{13}$C and $\delta^{15}$N values at each time point and location from the models described above (Magozzi, Somes). 
We then compared the simulated stable isotope profiles (Fig. S4) to the profile of the blue whale (Fig. 1).

@@

\section{Acknowledgments}
% adjectives not allowed!
This work was funded by the British Ecological Society (grant: 5771/6815). 
We thank XXX for comments on drafts.
We thank CJ Somes for providing  $\delta^{15}$N POM data.
